Laves polyhedra in synthetic tennantite, Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}, and its lattice dynamics

### 2021-02-03 SM/paper_tennantite.tex
proofreading
non highlighted text

\author[TISNCM]{Alexey~A.~Yaroslavzev}
\ead{yaroslavzevalex@gmail.com}
\author[MSU,ICRAS]{Alexey~N.~Kuznetsov}
\author[SIC]{Alexander~P.~Dudka}
\author[MSU]{Andrei~V.~Mironov}
\author[MIPT,TISNCM]{Sergey~G.~Buga}
\author[TISNCM]{Vladimir~V.~Denisov}

\address[TISNCM]{Technological Institute for Superhard and Novel Carbon Materials, 108840, Troitsk, Moscow, Russia}
\address[MSU]{Department of Chemistry, Lomonosov Moscow State University, 119991, Moscow, Russia}
\address[ICRAS]{Kurnakov Institute of General and Inorganic Chemistry RAS, 119991, Moscow, Russia}
\address[SIC]{Shubnikov Institute of Crystallography of Federal Scientific Research Centre “Crystallography and Photonics” of Russian Academy of Sciences, Leninskiy Prospekt 59, 119333, Moscow, Russia}
\address[MIPT]{Moscow Institute of Physics and Technology, 141700, 9 Institutsky lane, Dolgoprudny, Russia}

@@

The calculation of integral intensities from diffraction patterns was performed using the CrysAlisPro program\cite{Rigaku}.
The data processing included: consideration of geometric features of the data collection (Lorentz correction) and correction for polarization of radiation; calibration of the diffractometer\cite{Dudka2010}; absorption correction for \textcolor{red}{polyhedra} samples (performed by the Gaussian integration method\cite{Busing1957}); refinement of the half-wavelength contribution\cite{Dudka2010_2}; the correction for an extinction\cite{Becker1974}.
The corrections and refinement of structural parameters were performed in the Jana2006\cite{Petek2014} and ASTRA programs\cite{Dudka2007}.
Fourier syntheses of the electron density were calculated using the Jana2006 program\cite{Petek2014}.
Graphic constructions were made in Diamond\cite{Diamont}, and Vesta programs\cite{Momma2011}.

>>>>

\author[TISNCM]{Alexey~A.~Yaroslavzev}
\ead{yaroslavzevalex@gmail.com}
\author[MSU,ICRAS]{Alexey~N.~Kuznetsov}
\author[SIC]{Alexander~P.~Dudka}
\author[MSU]{Andrei~V.~Mironov}
\author[TISNCM]{Sergey~G.~Buga}
\author[TISNCM]{Vladimir~V.~Denisov}

\address[TISNCM]{Technological Institute for Superhard and Novel Carbon Materials, 108840, Troitsk, Moscow, Russia}
\address[MSU]{Department of Chemistry, Lomonosov Moscow State University, 119991, Moscow, Russia}
\address[ICRAS]{Kurnakov Institute of General and Inorganic Chemistry RAS, 119991, Moscow, Russia}
\address[SIC]{Shubnikov Institute of Crystallography of Federal Scientific Research Centre “Crystallography and Photonics” of Russian Academy of Sciences, Leninskiy Prospekt 59, 119333, Moscow, Russia}

@@

The calculation of integral intensities from diffraction patterns was performed using the CrysAlisPro program\cite{Rigaku}.
The data processing included: consideration of geometric features of the data collection (Lorentz correction) and correction for polarization of radiation; calibration of the diffractometer\cite{Dudka2010}; absorption correction for polyhedra samples (performed by the Gaussian integration method\cite{Busing1957}); refinement of the half-wavelength contribution\cite{Dudka2010_2}; the correction for an extinction\cite{Becker1974}.
The corrections and refinement of structural parameters were performed in the Jana2006\cite{Petek2014} and ASTRA programs\cite{Dudka2007}.
Fourier syntheses of the electron density were calculated using the Jana2006 program\cite{Petek2014}.
Graphic constructions were made in Diamond\cite{Diamont}, and Vesta programs\cite{Momma2011}.

### 2021-02-03 paper_tennantite0.tex
non highlighted text

\begin{abstract}
Synthetic tennantite, Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}, is the analogue of an abundant mineral that belongs to the tennantite-tetrahedri\textcolor{red}{t}e group with a low lattice thermal conductivity.
\textcolor{red}{Combined data from high-quality X-ray diffraction, electron microscopy (STEM--HAADF), Raman spectroscopy, as well as DFT calculations are used to analyze the peculiarities of its structure stability and dynamics.
Atomic displacement parameters (ADP) and low-energy optical phonon modes are discussed within the context of the structure variations and pecularities of the charge distribution.
The latter indicates that the tennantite structure tends to conform to the covalent polar bonds and its charge distribution is significantly affected by the atomic shifts in Laves polyhedra.
According to the DFT calculations, there is a population of stable model structures with varying shifts of copper atoms in Laves polyhedra, with total energies being very close (within 0.24~eV), which can explain the observed behavior of the ADP.
A specially designed  technique is used for the experimental analysis of the ADP that revealed Einstein characteristic temperatures in the tennantite structure to be in the range of 50--190~K, which is attributed to  low-energy optical phonon modes.}

\section*{Highlights}
\begin{itemize}
  \item Synthetic tennantite Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13} structure consists of disordered Laves \textcolor{red}{polyhedra}
  \item DFT calculations reveal a mixed valence of Cu atoms, and Laves \textcolor{red}{polyhedra} distortions
  \item Jahn-Teller distortions give rise to a set of unit cells with close \textcolor{red}{ground} state energies
  \item Rattling vibration modes refer to Einstein temperatures of 74, 104, 115, 185K
\end{itemize}

@@

The tennantite-tetrahedri\textcolor{red}{t}e group represents a large class of inorganic compounds, including complex copper sulfosalts with the general formula Cu\textsubscript{10+x}M\textsubscript{2-x}X\textsubscript{4}S\textsubscript{13}, where M is a transition metal (Zn, Fe, Hg, etc.), and X is a pnictogen (As, Sb, or Bi) \cite{Makovicky_2006}, with the pure tennantite itself represented by the formula Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}.
Despite being known for over two centuries, these minerals still represent objects of intense studies, both due to  structural peculiarities and potential applications. Arsenic- and antimony-containing compounds of this class are considered as advanced thermoelectric materials with a low lattice thermal conductivity\cite{Sootsman2009,Chetty2015}.
\textcolor{red}{For instance, tetrahedrite doped with Mn, Fe, Co, Ni, Zn, Te, Se showed ZT values in the range from 0.2 to 1.13\cite{Heo2014,Suekuni2013,Lu2012,Barbier2016,Rout2020,Zhu2019}.}

\textcolor{red}{The tennantite structure is treated as a sphalerite-type, thus conforms to the \hmn{I -4 3 m} space group\cite{yaroslavzev2019,Makovicky_2006}.
The structure can be described by five nonequivalent crystallographic sites: two for Cu, one for As, and two for S.
Cu1 is tetrahedrally coordinated (CuS\textsubscript{4}) by four S1, and As forms a trigonal pyramid (AsS\textsubscript{3}) with three S1.
Additionally, the S2 is a center of an octahedral unit that is called a Laves polyhedra (SCu\textsubscript{6}) with six copper atoms in sites Cu2 and Cu21.
Cu2 is trigonally coordinated (CuS\textsubscript{3}) by two S1 and one S2, while Cu21 is coordinated by two S1, one S2 and one As atoms.
Cu21 and Cu2 atoms show very large atomic displacements perpendicular to the plane of the triangle (Fig.~\ref{fig:generel_view}, right).
}
In a case of tennantite- and tetrahedrite-based materials, it is generally considered that a low lattice thermal conductivity is caused by hybridized Cu {\it 3d}  and chalcogen \textcolor{red}{{\it p}} orbitals\cite{Lai2015}.
Such a hybridization leads to a bonding asymmetry, and out-of-plane low frequency rattling modes, which are quasi-localized and anharmonic\cite{May2016,Bouyrie2015}.

\begin{figure}[ht]
\centering
\subfigure{\includegraphics[width=0.4\columnwidth]{general_view}}
 \quad
\subfigure{\includegraphics[width=0.35\columnwidth]{general_view_laves}}
\caption{\label{fig:generel_view} \textcolor{red}{Left --- the Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13} tennantite structure with coordination polyhedra around the Cu1 (blue), and the Laves polyhedra (yellow); right --- the Laves polyhedra containing Cu2 and Cu21 atoms with their partial occupancies shown as sections of spheres.} }
\end{figure}

@@

According to the structural study, these deviations take place due to the distortion of \textcolor{red}{Cu\textsubscript{6}S} octahedra \textcolor{red}{(Laves polyhedra)}\cite{Tanaka2015}.
The distortion at 84 K in tetrahedrite is reported to lead to either the symmetry lowering from \hmn{I -4 3 m} to \hmn{I -4 2 m}  space group accompanied by the displacement of S atoms\cite{Hathwar2019}, or to a shift of Cu2 atoms towards S\textsubscript{3} triangle plane with S\textsubscript{2} displacement\cite{Nasonova2016_2} in the same symmetry, and space group.

Recently, we have performed a multi-temperature (90--298~K) structural study of a synthetic tennantite\cite{yaroslavzev2019} and found that below ca. 120 K the copper atom distribution in tennantite changes from a more statistically delocalized Cu2 to a well-defined split into Cu2 and Cu21 positions without a first-order phase transition.
This effect is accompanied by As1 and S2 atoms displacement, which was also observed in \textcolor{red}{Ref.}~\cite{Hathwar2019}.
We also investigated the low- and room-temperature electronic structures of a synthetic tennantite based on the DFT calculations, which confirmed the second-order transition from the low-temperature AFM to the room-temperature PM state\cite{yaroslavzev2019}.

However, the relationships between the peculiarities of the tennantite crystal and its electronic structures and thermoelectric properties, including the behavior of Laves \textcolor{red}{polyhedra} in the structure, were put aside in the previous paper\cite{yaroslavzev2019}.
In this paper, we focus on these problems, and  investigate the tennantite cell energy distribution, and the charge density differences for different Cu positions in the Laves polyhedra.
Also, we investigate low-energy phonon modes, and peculiarities of the ADP in the temperature range of 85--293~K, the same as in Ref.~\cite{Dudka2019}, in the context of the cell energy distribution.
To clarify the distortions in Laves polyhedra, a study of the spherical single-crystal sample was performed and the high-angle annular dark-field (HAADF) images were obtained.

@@

\subsection{Single-crystal X-ray diffraction}\label{sec:level2}
A spherically shaped single-crystal sample of Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}  was prepared for the anisotropic extinction study\cite{Becker1975}.
Enraf-Nonius CAD-4 diffractometer with AgK\textsubscript{$\alpha$} equipped with graphite monochromator was used for the data collection.
The spherical single-crystal structural study results are presented in Supplementary Material, and they are compared with the structural data \textcolor{red}{from} the \textcolor{red}{regularly shaped crystal} sample\cite{yaroslavzev2019}.

@@

\subsection{Remarks on the structure of synthetic tennantite}\label{sec:level2}
The study of the high quality single crystal allowed us to re-investigate the crystal structure of Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}  with a high precision.
\textcolor{red}{The structure was refined in a way similar to previously described in Ref.~\cite{yaroslavzev2019}.
The sum of occupancies for Cu2, Cu21 and its symmetry related site were constrained to 1.
Due to rather large distance between these atoms their ADP were refined independently.
The maximum entropy method (MEM) refinement gave the result identical to \cite{yaroslavzev2019}.}
Our data confirms that the compound contains 12 copper atoms per formula unit, and shows no evidence of copper deficiency.
Neither XRD, nor TEM data \textcolor{red}{have shown detectable packing defects}.

@@

Due to the crystal growth method used, the sample shows a degree of anisotropic extinction. Extinction parameters are presented in the Table~\ref{tab:anis_ext}.
\textcolor{red}{The disorder of Cu21} is observed in the analyzed HAADF image of the sample in [110] zone (Fig.~\ref{fig:micro}).
The electron diffraction, and overlay of HAADF image with the tennantite crystal [110] zone generated by VESTA\cite{Momma2011}, are given in Supplementary Material.
Processed HAADF image (Fig.~\ref{fig:micro}, right) shows  an ellipticity for the Cu/As atomic columns, which is a measure of the atomic column elongation\cite{Nord2017}.
The atomic column elongation is evidenced by the blurring of the intensity distribution, and a large atomic displacement.
Similar effects were observed in tetrahedrite\cite{Mishra2017}, and several tetrahedral framework structures\cite{Suekuni2019}.
Also, the observed anisotropic displacement may possibly be caused by the crystal growth conditions or/and a crystal cell distortion.
However, in such case the structure must show prominent diagonal distortion and/or feature an increase or decrease of the copper content per cell (with respect to 12 copper atoms), which is not the case.

@@

Large static values for Cu2 atom (Table~\ref{tab:enst_temp}) \textcolor{red}{confirmed instability of this site\cite{Hathwar2019,yaroslavzev2019}}.
This disorder is reliably described by the anharmonic model of atomic displacements using the Gram-Charlier expansion to the 4th order.

@@

Within \textcolor{red}{the approach above}, the Einstein solid model is used to describe independent harmonic oscillators.
Thus, the low energy spectrum should consist of peaks with the Einstein's energies, which we found by DebyeFit (Table~\ref{tab:enst_temp}).
The results of the calculations are presented in Figure~\ref{fig:low_energ_raman}.
The Einstein's energies from structural study were used as the initial points for the fit, with the initial and fitted wavenumbers are presented and in Table~\ref{tab:ram_fit}.
The low energy $l_{1}$ and $l_{2}$ peaks are well described by Einstein's energies of Cu2, Cu21, Cu1, and As (66, 40, 79, and 122 cm\textsuperscript{-1} respectively).
\textcolor{red}{Same modes were observed and analyzed in tetrahedrites\cite{May2016,Lai2015} with energies approximately 4, 9 and 18 meV (33, 72, 145 cm\textsuperscript{-1}) which is
in good agreement with our results.
Thus, we can state that the reason for the strong anharmonic behavior is a dynamic disorder of the copper atoms and arsenic atoms.}
Consequently, this leads to out-of-plane low frequency rattling modes, and a low lattice thermal conductivity.

@@

\subsection{Theoretical analysis of possible polytypes and morphology of Laves \textcolor{red}{polyhedra} }\label{sec:level2}

Using the DFT calculations, we tested an assumption that the synthetic tennantite crystal structure is not the most energetically stable, and could  feature various polytypes.
As an initial point, we considered tennantite structure described in  \hmn{I -4 2 m}\ space group (henceforth called {\it reference structure}).
The unit cell contains Laves \textcolor{red}{polyhedra} formed by copper atoms around sulphur ones (Figure~\ref{fig:init_and_opt_laves}, number 0).
By shifting copper atoms by ca. 1~$\angstrom$, we produced a population of structures with varying shapes of Laves \textcolor{red}{polyhedra}, which were then subjected to the procedure of unconstrained structural optimization.
\textcolor{red}{In the DFT calculations, where we cannot have Cu2 and Cu21 at the same time since we cannot have part of an atom occupying a position within a single polyhedron, we prefer to only use Cu2 notation, but distinguish their positions before and after optimization.}
The Laves \textcolor{red}{polyhedra} morphologies before and after the optimization are shown in Figure~\ref{fig:init_and_opt_laves}.
Total energy differences between the reference structure of initial tennantite and optimized structures are shown in Figure~\ref{fig:energy_structure_rescale}.

\begin{figure}
\centering
 \includegraphics[width=0.48\textwidth]{Init_and_opt_renamed}
 \caption{\label{fig:init_and_opt_laves} The geometries of SCu\textsubscript{6} Laves \textcolor{red}{polyhedra} before and after structural optimization. sulphur atoms are shown in yellow, copper atoms before optimization are shown in grey, and after optimization in blue. VESTA\cite{Momma2011} was used for the visualisation. }
\end{figure}

@@

\begin{table*}
\caption{\label{tab:energy}%
The difference between the reference structure energy, and the unit cell energy of tennantite with different Laves \textcolor{red}{polyhedra}.
}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{ccccccccccc}
Structure, number & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\
The difference of unit cells energies, eV &  0.00 &  0.23 & -0.02 & -0.02 &  0.05 &  0.13 & -0.00 &  0.23 &  0.06 & -0.01 \\
 \\
Structure, number & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 \\
The difference of unit cells energies, eV &  0.16 & -0.02 &  0.05 &  0.03 & -0.02 &  0.02 &  0.12 & -0.02 & -0.03 & -0.04  \\
\end{tabular}}
\end{table*}

@@

\begin{figure}
\centering
 \includegraphics[width=0.48\textwidth]{energy_structure_rescale_ri}
 \caption{\label{fig:energy_structure_rescale} The difference between the reference structure energy and the unit cell energy of tennantite with different Laves \textcolor{red}{polyhedra}. Horizontal red line shows the reference unit cell energy. }
\end{figure}

Each of these groups has structures with an energy gain over the reference structure, and each of them has a certain split of a positive charge on Cu2 atoms, while Cu1 atomic charges remain constant throughout.
Thus, we can say that the splitting of the Cu2 position into Cu2 and Cu21, as evident from the XRD structure analysis, facilitates the differences in charge distribution.
Hence, the energy gain from forming Cu21 position leads to slightly different electronic states of Cu2 and Cu21 atoms, which, in turn, leads to the formation of asymmetric bonds, as was pointed out for tetrahedrites\cite{Lai2015,Long2020}.
We can also observe certain similarities between tennantite and inorganic compounds featuring atoms with  {\it 3d\textsuperscript{9}}  configuration, and having asymmetric Fermi surface.
\textcolor{red}{The resulting complex charge distribution, and variances in the copper atom charges and, therefore, effective oxidation states, caused by structural shifts, may, in turn, be the reason for complex exchange interactions and the reported AFM ordering\cite{yaroslavzev2019} in tennantite.}
Moreover, the same structure features have been described for tetrahedrite\cite{DiPaola2020}.
\textcolor{red}{The authors} noted, that understanding these features provides a key to descriptions of the phonon-limited electrical resistivity, and the Lorenz number.

It must be noted that the majority of the favorable structures belongs to group 2, i.e. more covalent structures with both arsenic and sulphur atoms showing relatively low positive and negative charges, respectively.
Interestingly, the reference structure is significantly more ionic in this respect.
Also, we must point out that, according to the DFT calculations, several structures have energy minimum within hundredths of eV (Table~\ref{tab:energy}), i.e. only several~kJ/mol (1~eV~$\sim$~96.5 kJ/mol), which is on the limit of the accuracy of the method.
This indicates, that the tennantite experimental structure may be \textcolor{red}{an averaged} result of several variations of the copper atoms shifts.

\section{Conclusions}\label{sec:level1}
In conclusion, using a combination of several experimental techniques, we have investigated the structural variations of synthetic tennantite, Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}, manifesting itself as a dynamic disorder of the copper atoms.
\textcolor{red}{The tennantite structure tends to conform to the asymmetric covalent polar bonds and its charge distribution is significantly affected by the atomic shifts in Laves polyhedra.
These effects, in turn, give rise to the out-of-plane low frequency rattling modes, well described by Einstein characteristic temperatures: 74, 104, 115, and 185~K.
The peculiarities of the tennantite crystal and electronic structure, including lattice dynamics and shifting charge distribution, shows similarities between this compound and inorganic compounds with an asymmetric Fermi surface, thus making tennantite--tetrahedrite group attractive research objects.}

>>>>

\begin{abstract}
Synthetic tennantite, Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}, is the analogue of an abundant mineral that belongs to the tennantite-tetrahedrite group with a low lattice thermal conductivity.
Combined data from high-quality X-ray diffraction, electron microscopy (STEM--HAADF), Raman spectroscopy, as well as DFT calculations are used to analyze the peculiarities of its structure stability and dynamics.
Atomic displacement parameters (ADP) and low-energy optical phonon modes are discussed within the context of the structure variations and pecularities of the charge distribution.
The latter indicates that the tennantite structure tends to conform to the covalent polar bonds and its charge distribution is significantly affected by the atomic shifts in Laves polyhedra.
According to the DFT calculations, there is a population of stable model structures with varying shifts of copper atoms in Laves polyhedra, with total energies being very close (within 0.24~eV), which can explain the observed behavior of the ADP.
A specially designed  technique is used for the experimental analysis of the ADP that revealed Einstein characteristic temperatures in the tennantite structure to be in the range of 50--190~K, which is attributed to  low-energy optical phonon modes.

\section*{Highlights}
\begin{itemize}
  \item Synthetic tennantite Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13} structure consists of disordered Laves polyhedra
  \item DFT calculations reveal a mixed valence of Cu atoms, and Laves polyhedra distortions
  \item Jahn-Teller distortions give rise to a set of unit cells with close ground state energies
  \item Rattling vibration modes refer to Einstein temperatures of 74, 104, 115, 185K
\end{itemize}

@@

The tennantite-tetrahedrite group represents a large class of inorganic compounds, including complex copper sulfosalts with the general formula Cu\textsubscript{10+x}M\textsubscript{2-x}X\textsubscript{4}S\textsubscript{13}, where M is a transition metal (Zn, Fe, Hg, etc.), and X is a pnictogen (As, Sb, or Bi) \cite{Makovicky_2006}, with the pure tennantite itself represented by the formula Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}.
Despite being known for over two centuries, these minerals still represent objects of intense studies, both due to  structural peculiarities and potential applications. Arsenic- and antimony-containing compounds of this class are considered as advanced thermoelectric materials with a low lattice thermal conductivity\cite{Sootsman2009,Chetty2015}.
For instance, tetrahedrite doped with Mn, Fe, Co, Ni, Zn, Te, Se showed ZT values in the range from 0.2 to 1.13\cite{Heo2014,Suekuni2013,Lu2012,Barbier2016,Rout2020,Zhu2019}.

The tennantite structure is treated as a sphalerite-type, thus conforms to the \hmn{I -4 3 m} space group\cite{yaroslavzev2019,Makovicky_2006}.
The structure can be described by five nonequivalent crystallographic sites: two for Cu, one for As, and two for S.
Cu1 is tetrahedrally coordinated (CuS\textsubscript{4}) by four S1, and As forms a trigonal pyramid (AsS\textsubscript{3}) with three S1.
Additionally, the S2 is a center of an octahedral unit that is called a Laves polyhedra (SCu\textsubscript{6}) with six copper atoms in sites Cu2 and Cu21.
Cu2 is trigonally coordinated (CuS\textsubscript{3}) by two S1 and one S2, while Cu21 is coordinated by two S1, one S2 and one As atoms.
Cu21 and Cu2 atoms show very large atomic displacements perpendicular to the plane of the triangle (Fig.~\ref{fig:generel_view}, right).

In a case of tennantite- and tetrahedrite-based materials, it is generally considered that a low lattice thermal conductivity is caused by hybridized Cu {\it 3d}  and chalcogen {\it p} orbitals\cite{Lai2015}.
Such a hybridization leads to a bonding asymmetry, and out-of-plane low frequency rattling modes, which are quasi-localized and anharmonic\cite{May2016,Bouyrie2015}.

\begin{figure}[ht]
\centering
\subfigure{\includegraphics[width=0.4\columnwidth]{general_view}}
 \quad
\subfigure{\includegraphics[width=0.35\columnwidth]{general_view_laves}}
\caption{\label{fig:generel_view} Left --- the Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13} tennantite structure with coordination polyhedra around the Cu1 (blue), and the Laves polyhedra (yellow); right --- the Laves polyhedra containing Cu2 and Cu21 atoms with their partial occupancies shown as sections of spheres.}
\end{figure}

@@

According to the structural study, these deviations take place due to the distortion of Cu\textsubscript{6}S octahedra (Laves polyhedra)\cite{Tanaka2015}.
The distortion at 84 K in tetrahedrite is reported to lead to either the symmetry lowering from \hmn{I -4 3 m} to \hmn{I -4 2 m}  space group accompanied by the displacement of S atoms\cite{Hathwar2019}, or to a shift of Cu2 atoms towards S\textsubscript{3} triangle plane with S\textsubscript{2} displacement\cite{Nasonova2016_2} in the same symmetry, and space group.

Recently, we have performed a multi-temperature (90--298~K) structural study of a synthetic tennantite\cite{yaroslavzev2019} and found that below ca. 120 K the copper atom distribution in tennantite changes from a more statistically delocalized Cu2 to a well-defined split into Cu2 and Cu21 positions without a first-order phase transition.
This effect is accompanied by As1 and S2 atoms displacement, which was also observed in Ref.~\cite{Hathwar2019}.
We also investigated the low- and room-temperature electronic structures of a synthetic tennantite based on the DFT calculations, which confirmed the second-order transition from the low-temperature AFM to the room-temperature PM state\cite{yaroslavzev2019}.

However, the relationships between the peculiarities of the tennantite crystal and its electronic structures and thermoelectric properties, including the behavior of Laves polyhedra in the structure, were put aside in the previous paper\cite{yaroslavzev2019}.
In this paper, we focus on these problems, and  investigate the tennantite cell energy distribution, and the charge density differences for different Cu positions in the Laves polyhedra.
Also, we investigate low-energy phonon modes, and peculiarities of the ADP in the temperature range of 85--293~K, the same as in Ref.~\cite{Dudka2019}, in the context of the cell energy distribution.
To clarify the distortions in Laves polyhedra, a study of the spherical single-crystal sample was performed and the high-angle annular dark-field (HAADF) images were obtained.

@@

\subsection{Single-crystal X-ray diffraction}\label{sec:level2}
A spherically shaped single-crystal sample of Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}  was prepared for the anisotropic extinction study\cite{Becker1975}.
Enraf-Nonius CAD-4 diffractometer with AgK\textsubscript{$\alpha$} equipped with graphite monochromator was used for the data collection.
The spherical single-crystal structural study results are presented in Supplementary Material, and they are compared with the structural data from the regularly shaped crystal sample\cite{yaroslavzev2019}.

@@

\subsection{Remarks on the structure of synthetic tennantite}\label{sec:level2}
The study of the high quality single crystal allowed us to re-investigate the crystal structure of Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}  with a high precision.
The structure was refined in a way similar to previously described in Ref.~\cite{yaroslavzev2019}.
The sum of occupancies for Cu2, Cu21 and its symmetry related site were constrained to 1.
Due to rather large distance between these atoms their ADP were refined independently.
The maximum entropy method (MEM) refinement gave the result identical to \cite{yaroslavzev2019}.
Our data confirms that the compound contains 12 copper atoms per formula unit, and shows no evidence of copper deficiency.
Neither XRD, nor TEM data have shown detectable packing defects.

@@

Due to the crystal growth method used, the sample shows a degree of anisotropic extinction. Extinction parameters are presented in the Table~\ref{tab:anis_ext}.
The disorder of Cu21 is observed in the analyzed HAADF image of the sample in [110] zone (Fig.~\ref{fig:micro}).
The electron diffraction, and overlay of HAADF image with the tennantite crystal [110] zone generated by VESTA\cite{Momma2011}, are given in Supplementary Material.
Processed HAADF image (Fig.~\ref{fig:micro}, right) shows  an ellipticity for the Cu/As atomic columns, which is a measure of the atomic column elongation\cite{Nord2017}.
The atomic column elongation is evidenced by the blurring of the intensity distribution, and a large atomic displacement.
Similar effects were observed in tetrahedrite\cite{Mishra2017}, and several tetrahedral framework structures\cite{Suekuni2019}.
Also, the observed anisotropic displacement may possibly be caused by the crystal growth conditions or/and a crystal cell distortion.
However, in such case the structure must show prominent diagonal distortion and/or feature an increase or decrease of the copper content per cell (with respect to 12 copper atoms), which is not the case.

@@

Large static values for Cu2 atom (Table~\ref{tab:enst_temp}) confirmed instability of this site\cite{Hathwar2019,yaroslavzev2019}.
This disorder is reliably described by the anharmonic model of atomic displacements using the Gram-Charlier expansion to the 4th order.

@@

Within the approach above, the Einstein solid model is used to describe independent harmonic oscillators.
Thus, the low energy spectrum should consist of peaks with the Einstein's energies, which we found by DebyeFit (Table~\ref{tab:enst_temp}).
The results of the calculations are presented in Figure~\ref{fig:low_energ_raman}.
The Einstein's energies from structural study were used as the initial points for the fit, with the initial and fitted wavenumbers are presented and in Table~\ref{tab:ram_fit}.
The low energy $l_{1}$ and $l_{2}$ peaks are well described by Einstein's energies of Cu2, Cu21, Cu1, and As (66, 40, 79, and 122 cm\textsuperscript{-1} respectively).
Same modes were observed and analyzed in tetrahedrites\cite{May2016,Lai2015} with energies approximately 4, 9 and 18 meV (33, 72, 145 cm\textsuperscript{-1}) which is
in good agreement with our results.
Thus, we can state that the reason for the strong anharmonic behavior is a dynamic disorder of the copper atoms and arsenic atoms.
Consequently, this leads to out-of-plane low frequency rattling modes, and a low lattice thermal conductivity.

@@

\subsection{Theoretical analysis of possible polytypes and morphology of Laves polyhedra}\label{sec:level2}

Using the DFT calculations, we tested an assumption that the synthetic tennantite crystal structure is not the most energetically stable, and could  feature various polytypes.
As an initial point, we considered tennantite structure described in  \hmn{I -4 2 m}\ space group (henceforth called {\it reference structure}).
The unit cell contains Laves polyhedra formed by copper atoms around sulphur ones (Figure~\ref{fig:init_and_opt_laves}, number 0).
By shifting copper atoms by ca. 1~$\angstrom$, we produced a population of structures with varying shapes of Laves polyhedra, which were then subjected to the procedure of unconstrained structural optimization.
In the DFT calculations, where we cannot have Cu2 and Cu21 at the same time since we cannot have part of an atom occupying a position within a single polyhedron, we prefer to only use Cu2 notation, but distinguish their positions before and after optimization.
The Laves polyhedra morphologies before and after the optimization are shown in Figure~\ref{fig:init_and_opt_laves}.
Total energy differences between the reference structure of initial tennantite and optimized structures are shown in Figure~\ref{fig:energy_structure_rescale}.

\begin{figure}
\centering
 \includegraphics[width=0.48\textwidth]{Init_and_opt_renamed}
 \caption{\label{fig:init_and_opt_laves} The geometries of SCu\textsubscript{6} Laves polyhedra before and after structural optimization. sulphur atoms are shown in yellow, copper atoms before optimization are shown in grey, and after optimization in blue. VESTA\cite{Momma2011} was used for the visualisation. }
\end{figure}

@@

\begin{table*}
\caption{\label{tab:energy}%
The difference between the reference structure energy, and the unit cell energy of tennantite with different Laves polyhedra.
}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{ccccccccccc}
Structure, number & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\
The difference of unit cells energies, eV &  0.00 &  0.23 & -0.02 & -0.02 &  0.05 &  0.13 & -0.00 &  0.23 &  0.06 & -0.01 \\
 \\
Structure, number & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 \\
The difference of unit cells energies, eV &  0.16 & -0.02 &  0.05 &  0.03 & -0.02 &  0.02 &  0.12 & -0.02 & -0.03 & -0.04  \\
\end{tabular}}
\end{table*}

@@

\begin{figure}
\centering
 \includegraphics[width=0.48\textwidth]{energy_structure_rescale_ri}
 \caption{\label{fig:energy_structure_rescale} The difference between the reference structure energy and the unit cell energy of tennantite with different Laves polyhedra. Horizontal red line shows the reference unit cell energy. }
\end{figure}

Each of these groups has structures with an energy gain over the reference structure, and each of them has a certain split of a positive charge on Cu2 atoms, while Cu1 atomic charges remain constant throughout.
Thus, we can say that the splitting of the Cu2 position into Cu2 and Cu21, as evident from the XRD structure analysis, facilitates the differences in charge distribution.
Hence, the energy gain from forming Cu21 position leads to slightly different electronic states of Cu2 and Cu21 atoms, which, in turn, leads to the formation of asymmetric bonds, as was pointed out for tetrahedrites\cite{Lai2015,Long2020}.
We can also observe certain similarities between tennantite and inorganic compounds featuring atoms with  {\it 3d\textsuperscript{9}}  configuration, and having asymmetric Fermi surface.
The resulting complex charge distribution, and variances in the copper atom charges and, therefore, effective oxidation states, caused by structural shifts, may, in turn, be the reason for complex exchange interactions and the reported AFM ordering\cite{yaroslavzev2019} in tennantite.
Moreover, the same structure features have been described for tetrahedrite\cite{DiPaola2020}.
The authors noted, that understanding these features provides a key to descriptions of the phonon-limited electrical resistivity, and the Lorenz number.

It must be noted that the majority of the favorable structures belongs to group 2, i.e. more covalent structures with both arsenic and sulphur atoms showing relatively low positive and negative charges, respectively.
Interestingly, the reference structure is significantly more ionic in this respect.
Also, we must point out that, according to the DFT calculations, several structures have energy minimum within hundredths of eV (Table~\ref{tab:energy}), i.e. only several~kJ/mol (1~eV~$\sim$~96.5 kJ/mol), which is on the limit of the accuracy of the method.
This indicates, that the tennantite experimental structure may be an averaged result of several variations of the copper atoms shifts.

\section{Conclusions}\label{sec:level1}
In conclusion, using a combination of several experimental techniques, we have investigated the structural variations of synthetic tennantite, Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}, manifesting itself as a dynamic disorder of the copper atoms.
The tennantite structure tends to conform to the asymmetric covalent polar bonds and its charge distribution is significantly affected by the atomic shifts in Laves polyhedra.
These effects, in turn, give rise to the out-of-plane low frequency rattling modes, well described by Einstein characteristic temperatures: 74, 104, 115, and 185~K.
The peculiarities of the tennantite crystal and electronic structure, including lattice dynamics and shifting charge distribution, shows similarities between this compound and inorganic compounds with an asymmetric Fermi surface, thus making tennantite--tetrahedrite group attractive research objects.
